Moment Distances from Robust Subspace for Botnet Detection

### 2020-04-28 docs/papers/2019JNCA_m-rpca/main.tex
updating JNCA paper

\chapter{Robust Framework for Detection of Network Attacks in Imbalanced Traffic}
\label{ch:4_m_rpca}

According to \cite{distilnetworks2019}, 37.9\% of all Internet traffic of 2018 was not from human activities, but from bots, that can be classified as good bots, that perform legitimate operations, and bad bots, which perform malicious activities, such as DDoS attacks, probe attack or frauds. In 2018, bad bots accounted for 20.4\% of all website traffic.

According to \cite{BISSELL2019} the proportion of cost on discovery activities for cybersecurity has increased steadily since 2015. Forensic cyber and user behavior analytics also present an opportunity for cost savings — US\$1.72 million — with discovery and investigation activities. However, only 32\% of organizations have deployed these technologies enterprise-wide. Accenture consulting argues that 89\% of survey respondents believe breakthrough technologies, like machine learning and user behavior analytics, are essential to securing the future of their organizations \cite{Accenture2018}.

To face the adversarial model, network attacks and counter measures of attackers to avoid detection, it is possible to adopt unsupervised or semi-supervised approaches for network anomaly detection, by means of behavioral analysis, where it is not necessary known anomalies for training models \cite{moustafa2019holistic}.

Anomalies in the context of network traffic can be hard to identify and separate from legitimate data due to the rare occurrences of anomalies in comparison to legitimate events. Therefore, anomaly detection algorithms have to be highly discriminating, robust to corruption and able to deal with the imbalanced data problem \cite{he2008learning}. Note that data corruption refers to outliers that can be part of the data but not be considered malicious. The imbalanced data problem corresponds to data sets exhibiting significant imbalances of classes or rare events of some classes \cite{he2009learning}, which can be legitimate or malicious classes in network anomaly detection problems.

Traditionally adopted algorithms for anomaly detection assume a Gaussian or symmetric distributed data \cite{lakhina2005mining}, however this assumption may not be observed in some real world problems \cite{benson2010network}, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{leon2017probability}.

@@

The findings presented by Benson \emph{et al.} \cite{benson2010network} and Leon-Garcia \cite{leon2017probability} show that the skewness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian distributed data for network anomaly detection. Additionally, the fitting of network traffic to skewed and heavy-tailed distributions can indicate opportunities to exploit properties and characteristics of the skewness and heavy-tailed distributions to obtain improved classifiers for network anomaly detection.

Network anomaly detection problems are usually characterized by imbalanced data \cite{Phua2004minority}. However, learning algorithms for imbalanced data has been a challenging research topic, considering that the imbalanced data can compromise the performance of most standard learning algorithms, creating bias or unfair weight to learn from the majority class and reducing detection capacity of anomalies that are characterized by the minority class \cite{he2008learning}. Therefore, learning methods for imbalanced and skewed data have attracted attention of researchers \cite{hubert2009robustskewed}.

Robust subspace learning has been attracting a growing attention of researchers aiming the development of network attack detection systems \cite{pascoal2012robust, zhou2017anomaly} that rely on behavioral analysis. An outlyingness-approach based on a robust estimator of skewness, combined with robust estimators of location and scale, can be able to flag the outlying measurements \cite{hubert2009robustskewed}. According to Hubert \emph{et al.} \cite{hubert2009robustskewed}, when the same methodology would be used with non-robust estimators of location, scale and skewness, the outlyingness-values would be affected by the outliers such that the outlying group could be masked.

Considering that the skewness of anomalous and legitimate traffic can highlight features for improving anomaly detection and network attack detection in imbalanced data, and considering that the distance between robust estimates can be used for network attack detection, we propose the Moment-based Robust Principal Component Analysis (m-RPCA), which is an approach based on distances between moments computed from a robust subspace learned by Robust Principal Component Analysis (RPCA) from contaminated observations, in order to detect anomalies from skewed data and network traffic.

The proposed approach relies on a robust subspace computed from supposed legitimate observations, for estimating the moments to be used for distance analysis. The anomaly detection from contaminated observations evaluate the Mahalanobis distance between the robust moments and new contaminated observations, in a semi-supervised fashion, without the computational cost of new robust subspace learning for anomaly detection from new observations. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

We evaluate the accuracy of the m-RPCA for anomaly detection on simulated data sets, with skewed and heavy-tailed distributions, and for network attack detection on CTU-13 data set \cite{garcia2014empirical}, which is a large data set of legitimate, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}. The Experimental evaluation compares our proposal to standard and widely adopted algorithms for anomaly detection, which are based on clustering and statistical approaches, and to ROBPCA \cite{hubert2005robpca}, which is a anomaly detection method that relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contribution of this work is the proposal of a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results showing improvements through an experimental evaluation on simulated skewed and heavy-tailed distributions and on real data set with legitimate network traffic and botnet attacks.

This chapter is organized as follows. In Section \ref{sec:3_relatedworks}, a literature review about network anomaly detection, botnet detection, and imbalanced learning is conducted. We present in Section \ref{sec:3_datamodel} the data model and the evaluated data set. In Section \ref{sec:3_proposal} it is described the proposed approach for network attack detection. We discuss in Section \ref{sec:3_experiments} the experimental validation, and in Section \ref{sec:3_simulated_result} are presented the results of the simulated scenarios and in Section \ref{sec:3_ctu13_result} we describe the results for anomaly detection on CTU-13 data set. Finally, in Section \ref{sec:3_conclusion} we draw the conclusions and the suggestions for future work.

\section{Related Works}
\label{sec:3_relatedworks}

% network anomaly detection
Network Anomaly detection has emerging as an important approach for securing communication networks and to deal with the increasing number of network attacks. Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques, which include classification, statistical, information theory and clustering. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), describing details of cyber-attacks and new solutions for anomaly detection, and provides a benchmark data sets for training and validating approaches for network anomaly detection.

% botnet detection
A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information ex-filtration. We refer to \cite{ahmed2016survey} and \cite{moustafa2019holistic} for an overview of network attacks. The botmaster coordinate a botnet through a command and control (C\&C) channel where bots receive commands and synchronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and architectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

According to \cite{distilnetworks2019}, 37.9\% of all Internet traffic of 2018 was not from human activities, but from bots, that can be classified as good bots, that perform legitimate operations, and bad bots, which are responsible by malicious activities, such as DDoS, probe attack or frauds.

@@

Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that removed the redundant data records, in order to avoid biased classifications. However, NSL-KDD data set maintain the limitations of the KDD 99 regarding volume and lacks on reproduction of recent network traffic and attacks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods by means of a simple and reproducible methodology, by a good data set and by a new error metric. This paper evaluates some data sets for network anomaly detection, and surveys some approaches for botnet detection, and proposes two methods (BClus and CAMNEP) for botnet detection, comparing results to BotHunter \cite{gu2007bothunter}.

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, legitimate and background labeled data, in an imbalanced distribution like in a real networks. The authors recommend scenarios for training and testing in order to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Taking into account the adoption of unsupervised or semi-supervised approaches for anomaly detection, adopting the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

Wang and Paschalidis \cite{wang2017botnet} proposed a botnet detection approach based on anomaly and community detection, aiming for detecting botnets and identifying bots before the botnet becomes active. The first stage detects anomalies by leveraging large deviations of an empirical distribution. The second stage detects the bots using ideas from social network community detection in a graph that captures correlations of interactions among nodes over time. This work is compared to the BotHunter \cite{gu2007bothunter} for the CTU-13 botnet data set \cite{garcia2014empirical}.

% Anomaly detection by subspace learning
Traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original feature space \cite{ringberg2007sensitivity}. However, to overcome the above mentioned limitations, some approaches based on PCA have been proposed for network anomaly detection.

Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the network traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of legitimate and anomaly observations according to principal components (legitimate) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \citeme{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

@@

RPCA has been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Hubert \emph{et al.} \cite{hubert2009robustskewed} evaluated ROBPCA-AO for real and simulated data sets, but it is not clear if this method was evaluated for very skewed and imbalanced data sets, such as in the network attack detection problem.

Robust subspace learning has received a growing attention of researchers aiming the development of network anomaly detection systems \cite{ hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly.

Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA approach, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}.

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. MD is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\pmb{\mu}$ and covariance $\hat{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the Mahalanobis Distance between robust estimates and contaminated observations can reveal anomalies.

We propose an approach based on the distances between moments computed from learned robust subspace and contaminated observations, for anomaly detection on skewed and imbalanced data sets, and evaluate our approach for network attack detection. The proposed approach relies on learning robust subspace from supposed legitimate traffic for estimating the moments (mean, skewness and kurtosis).

Thus, the anomaly detection for contaminated observations should evaluate the distance between the robust moments and contaminated observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations, or in an unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is between the robust moments and the contaminated observations.

\section{Data Model}
\label{sec:3_datamodel}

In this chapter, we adopt the data model introduced in Section \ref{sec:2_datamodel} and define that the Frobenius norm is denoted as $\left\| \mathord{\cdot} \right\|_F$, while $\left\| \mathord{\cdot} \right\|_*$ denotes the nuclear norm of a matrix and $\left\| \mathord{\cdot} \right\|_1$ means the sum of the absolute values of matrix entries. We also define the operator $\langle \mathord{\cdot} \rangle$, that is the standard trace inner product, and the operator $[ \mathord{\cdot} ]^c$, which denotes the indexes of the $c$ largest values of a vector.

This section also presents a description of the simulated data model in Subsection \ref{sec:3_simulation} and in Subsection \ref{sec:3_CTU-13} we describe the data model for the CTU-13 data set.

>>>>

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{natbib}
\usepackage{supertabular}
\usepackage{fancybox}
\usepackage{acronym}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{setspace}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{listings}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{courier}
\usepackage{amsfonts}
\usepackage{morefloats}
\usepackage{lipsum}
\usepackage{mathtools}
\usepackage{amsthm,amsmath}

% table
\usepackage{makecell}
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

% appendix
\usepackage[titletoc,title]{appendix}
\usepackage{hyperref}
\usepackage{cleveref}

% norm

\modulolinenumbers[5]

\journal{Journal of Network and Computer Applications}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{Moment Distances from Robust Subspace for Botnet Detection}

%% Group authors per affiliation:
\author[unbaddress]{Thiago P. B. Vieira}
\author[unbaddress]{Eduardo S. C. Vilaça}
\author[unbaddress,Ilmenauaddress,Fraunhoferaddress]{João Paulo C. L. da Costa}
\author[unbaddress]{Éder S. Gualberto}
\author[unbaddress]{Rafael T. de Sousa Júnior}

\address[unbaddress]{Department of Electrical Engineering, University of Brasilia (UnB), 70910-900, Brasília-DF, Brazil}
\address[Ilmenauaddress]{Institute for Information Technology, Ilmenau University of Technology, Ilmenau, Germany}
\address[Fraunhoferaddress]{Fraunhofer Institute for Integrated Circuits IIS, Erlangen, Germany}

\begin{abstract}
The U.S. government estimates that malicious cyber activity cost the U.S. economy between US\$57 billion and US\$109 billion in 2016. Accenture argues that 89\% of survey respondents believe breakthrough technologies, like artificial intelligence, machine learning and user behavior analytics, are essential for securing organizations. The imbalanced data can compromise the performance of standard learning algorithms, creating bias or unfair weight to learn from the majority class and reducing detection capacity of anomalies. Some widely adopted algorithms for anomaly detection assume a Gaussian distributed data for legitimate observations, but this assumption may not be observed in network traffic, which is usually characterized by skewed and heavy-tailed distributions. We propose the Moment-based Robust Principal Component Analysis (m-RPCA) for botnet attack detection, which is a framework based on distances between contaminated observations and moments computed from a robust subspace learned by RPCA, to detect anomalies and network attacks. We evaluate m-RPCA on simulated data and on CTU-13 data set, and results show that m-RPCA can improve the botnet attack detection and the anomaly detection on skewed data, in comparison to anomaly detection algorithms.
\end{abstract}

\begin{keyword}
Anomaly Detection, Botnet Attack Detection, Skewed Data, Imbalanced Data, Robust Principal Component Analysis (RPCA).
\end{keyword}

\end{frontmatter}

\linenumbers

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}

According to \cite{distilnetworks2019}, 37.9\% of all Internet traffic of 2018 was not from human activities, but from bots, that can be classified as good or bad bots, which accounted for 20.4\% of all website traffic. A botnet is a collection of bots or systems for executing automated tasks, often in the form of compromised hosts, including desktops, mobile devices or things of the Internet of Things (IoT). Some examples of botnets include Cyclone, Mirai, Nitol and Sentry MBA \cite{Catucci2019}.

According to Hevesi \cite{Hevesi2019} in a Gartner's publication, botnets scan the Internet looking for unprotected IoT devices, which are projected to be 25 billion devices in 2021. Hevesi argues that if 1\% of IoT devices is added to a botnet and had access to 5G, the potential throughput would be 50 Gbps in a flooding attack.

Distributed attacks organized by botnet has increased and demanded the development of counter measures to detect and avoid unknown attacks or even to deal with adversarial changes of behavior, location and other patterns \cite{wang2017botnet, Wang2018ddosbotnetssurvey}. To face the adversarial model, network attacks and counter measures of attackers to avoid detection, it is possible to adopt unsupervised or semi-supervised approaches for network anomaly detection, by means of behavioral analysis, where it is not necessary known anomalies for training models \cite{moustafa2019holistic}.

Anomalies in the context of network traffic can be hard to identify and separate from legitimate data due to the rare occurrences of anomalies in comparison to legitimate events. Additionally, widely adopted algorithms for anomaly detection assume a Gaussian or symmetric distributed data \cite{lakhina2005mining}, however this assumption may not be observed in some real world problems \cite{benson2010network}, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{leon2017probability}.

@@

The findings presented by Benson \emph{et al.} \cite{benson2010network} and Leon-Garcia \cite{leon2017probability} show that the skewness and heavy-tailed distributions may be important for network traffic analysis, and can motivate researches to evaluate the impact of skewed data into algorithms that rely on Gaussian data for network anomaly detection. Additionally, the fitting of network traffic to skewed and heavy-tailed distributions can indicate opportunities to exploit characteristics of the skewness and heavy-tailed distributions to improve classifiers for network anomaly detection.

Network anomaly detection problems are usually characterized by imbalanced data \cite{Phua2004minority}. However, learning algorithms for imbalanced data has been a challenging topic, considering that imbalanced data can compromise the performance of most standard learning algorithms, creating bias or unfair weight to learn from the majority class and reducing the detection capacity \cite{he2008learning}. Therefore, learning methods for imbalanced and skewed data have attracted attention of researchers \cite{hubert2009robustskewed}.

Robust subspace learning has been attracting a growing attention of researchers aiming the development of network attack detection systems that rely on behavioral analysis \cite{pascoal2012robust, zhou2017anomaly}. An outlyingness-approach based on a robust estimator of skewness, combined with robust estimators of location and scale, can be able to flag the outlying measurements \cite{hubert2009robustskewed}. According to Hubert \emph{et al.} \cite{hubert2009robustskewed}, when the same methodology would be used with non-robust estimators of location, scale and skewness, the outlyingness-values would be affected by the outliers such that the outlying group could be masked.

Considering that the skewness of anomalous and legitimate traffic can highlight features for improving anomaly detection and network attack detection in imbalanced data, and considering that the distance between robust estimates can be used for network attack detection, we propose the Moment-based Robust Principal Component Analysis (m-RPCA), which is a framework based on distances between moments computed from a robust subspace learned by Robust Principal Component Analysis (RPCA) from contaminated observations, to detect anomalies from skewed data and network traffic.

The m-RPCA relies on a robust subspace computed from supposed legitimate observations, for estimating the moments to be used for distance analysis. The anomaly detection from contaminated observations evaluate the Mahalanobis distance between the robust moments and new contaminated observations, in a semi-supervised fashion. The m-RPCA can also be computed as an unsupervised algorithm, with subspace learning from the same contaminated data that is the target of the anomaly detection analysis.

We evaluate the accuracy of the m-RPCA for anomaly detection on simulated data set, with skewed and heavy-tailed distributions, and for botnet attack detection on CTU-13 data set \cite{garcia2014empirical}, which is a large data set of legitimate, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for network attack detection \cite{osanaiye2016distributed}. The experimental evaluation compares m-RPCA to standard and widely adopted algorithms for anomaly detection, which are based on clustering and statistical approaches, and to ROBPCA-AO \cite{hubert2005robpca}, which is an anomaly detection method that relies on robust estimates with adjusted outlyingness based on robust skewness.

The main contributions of this work are a novel semi-supervised and unsupervised method for anomaly detection in skewed and imbalanced data, with results showing improvements in experimental evaluation on simulated skewed and heavy-tailed data and on real data set with legitimate and botnet traffic.

This paper is organized as follows. In Section \ref{sec:review}, a literature review about network anomaly detection, botnet detection, and imbalanced learning is conducted. We present in Section \ref{sec:datamodel} the data model and the evaluated data set. In Section \ref{sec:mdrs} it is described the proposed framework for network attack detection. In Section \ref{sec:experimentalresults} we discuss the experimental validation, the results for simulated scenarios and the results for botnet attack detection on CTU-13 data set. Finally, in Section \ref{sec:conclusionandfutureworks} we draw the conclusions and the suggestions for future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% network anomaly detectiom, botnet detection, botnet detection approaches, approaches based on subspace learning, feature generation and aggregation, about the methodos to compare and metrics
\section{Literature Review}
\label{sec:review}

% network anomaly detection
Network Anomaly detection has emerging as an important approach for securing communication networks and to deal with the increasing number of network attacks. Bhuyan \emph{et al.} \cite{bhuyan2014network} provide an overview of facets of network anomaly detection, present attacks normally encountered by network intrusion detection systems, and categorize existing network anomaly detection methods and systems based on the underlying techniques. Ahmed \emph{et al.} \cite{ahmed2016survey} present an analysis of four major categories of anomaly detection techniques, which include classification, statistical, information theory and clustering approaches. Moustafa \emph{et al.} \cite{moustafa2019holistic} discuss aspects of anomaly-based Network Intrusion Detection Systems (NIDSs), describing details of cyber-attacks and new solutions for anomaly detection, and provides a benchmark data set for training and validating approaches for network anomaly detection.

% botnet detection
According to \cite{distilnetworks2019}, 37.9\% of all Internet traffic of 2018 was not from human activities, but from bots, that can be classified as good bots or bad bots. A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities, such as spam, phishing, identity theft, and information ex-filtration. We refer to \cite{ahmed2016survey} and \cite{moustafa2019holistic} for an overview of network attacks. The botmaster coordinate a botnet through a command and control (C\&C) channel where bots receive commands and synchronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and architectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

@@

Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data set for accuracy and performance evaluation, due to their public availability and labeled traffic. Even though the KDD 99 are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data set in use by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 that removed the redundant data records, to avoid biased classifications. However, NSL-KDD maintain the limitations of the KDD 99 regarding volume and lacks on reproduction of recent network traffic and attacks.

Garcia \emph{et al.} \cite{garcia2014empirical} compare three botnet detection methods by means of a simple and reproducible methodology, by a good data set and by a new error metric. This paper evaluates data sets for network anomaly detection, and surveys approaches for botnet detection, and proposes two methods (BClus and CAMNEP) for botnet detection, comparing results to BotHunter \cite{gu2007bothunter}.

Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by attack, legitimate and background labeled data, in an imbalanced distribution and according to real network traffics. The authors recommend scenarios for training and testing to avoid the use of traffic from a botnet family for training and testing, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Taking into account the adoption of unsupervised or semi-supervised approaches for anomaly detection, if adopted the training and testing approach proposed by Garcia \emph{et al.}, some botnet malwares wouldn't be tested, since in the author's proposal, some botnets are present only for training.

Wang and Paschalidis \cite{wang2017botnet} proposed a botnet detection approach based on anomaly and community detection, aiming for detecting botnets and identifying bots before the botnet becomes active. The first stage detects anomalies by leveraging large deviations of an empirical distribution. The second stage detects the bots using ideas from social network community detection in a graph that captures correlations of interactions among nodes over time. This work is compared to the BotHunter \cite{gu2007bothunter} for the CTU-13 data set \cite{garcia2014empirical}.

% Anomaly detection by subspace learning
According to Ringberg \emph{et al.} \cite{ringberg2007sensitivity}, traditional PCA-based anomaly detection models are not suitable for anomaly interpretation, as they judge whether a data instance is an anomaly or not based on the length of its projection on the abnormal subspace spanned by the less significant principal components, and there is no direct mapping between PCA’s dimensionality-reduced subspace and the original subspace \cite{ringberg2007sensitivity}. To overcome the above mentioned limitations, some approaches based on PCA have been proposed for network anomaly detection.

Callegari \emph{et al.} \cite{callegari2011novel} proposed a PCA-based method for identifying the network traffic flows responsible for an anomaly detected at the aggregate level, by means of a separation of legitimate and anomaly observations according to principal components (legitimate) and remaining (anomalies). Lee \emph{et al.} \cite{Lee2013} presented OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. Vieira \emph{et al.} \cite{vieira2017model} proposed a framework that applies Model Order Selection (MOS) for detection of time frames under attack and uses similarity analysis to extract details and detect the time and ports under attack.

@@

RPCA has been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA-AO \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of mean and covariance matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Hubert \emph{et al.} \cite{hubert2009robustskewed} evaluated ROBPCA-AO for real and simulated data set, but it is not clear if this method was evaluated for very skewed and imbalanced data set, such as in the network attack detection problem.

Robust subspace learning has received a growing attention of researchers aiming the development of network anomaly detection systems \cite{ hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} \cite{pascoal2012robust} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly.

Zhou and Paffenroth \cite{zhou2017anomaly} proposed the Robust Deep Autoencoders (RDA), which the central idea is that a RDA inherits the non-linear representation capabilities of autoencoders combined with the anomaly detection capabilities of RPCA. Considering that outliers and noise may reduce the quality of representations discovered by deep autoencoders, the proposed model isolates noise and outliers in the input by means of a RPCA, and the autoencoder is trained after this isolation. RDA was evaluated by the authors for the MNIST data set.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}.

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples, by considering the covariance between the variables and their mean values. The MD is a measure of the distance between a vector $\pmb{x}$ and a distribution $\pmb{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936md}. MD is a multi-dimensional generalization for measuring how many standard deviations away $\pmb{x}$ is from the mean $\pmb{\mu}$ and covariance $\hat{\pmb{\Sigma}}$ of $\pmb{X}$. MD has been used for distance based anomaly detection with robust estimates in many areas, assuming that the MD between robust estimates and contaminated observations can reveal anomalies.

We propose a framework based on the distances between moments computed from learned robust subspace and contaminated observations, for anomaly detection on skewed and imbalanced data, and for botnet attack detection.

The proposed semi-supervised approach relies on learning robust subspace from supposed legitimate traffic for estimating the moments (mean, skewness and kurtosis), and the anomaly detection is computed from contaminated observations by means of the distance between the robust moments and contaminated observations, without the computational cost of new robust subspace learning to detect anomalies on new observations. The m-RPCA can also be adopted as an unsupervised approach, where the robust subspace is learned from the contaminated data and the distance analysis is computed between the robust moments and the contaminated observations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Data Model}
\label{sec:datamodel}
